---
title: "coursera_project1"
author: "Visha"
date: "16/12/2023"
output: html_document
---

Setting up the chunks to print the code to the document as the default by setting knitr::opts_chunk$set(echo = TRUE)
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown script for Course Project 1- Reproducible Research


```{r libraries}
# load the libraries required for the analysis
library(ggplot2)
library(dittoSeq)
library(data.table)
library(magrittr)
library(readxl)
library(dplyr)
```

# This chunk reads in the data and gives a brief summary of the raw and unprocessed data

```{r load_data, echo=TRUE}
# read in the csv file and store it in a R variable called activity
activity = read.csv("~/Downloads/activity.csv")

# take a look at the unprocessed data
head(activity)

# summary of the activity 
summary(activity)

# calculate the missing values in the data
missing_values <- sapply(activity, function(x) sum(is.na(x)))

# Display the results
print("Total number of missing values per column:")
print(missing_values)

```

# This chunk renders the Histogram of the total number of steps taken each day

```{r histogram, echo=TRUE}
# Convert the 'date' column to a Date type
activity$date <- as.Date(activity$date)

# Calculate the total number of steps taken each day
total_steps_per_day <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)

# take a look at the total number of steps taken each day
head(total_steps_per_day)

# Create a histogram
hist(total_steps_per_day$steps, 
     main = "Histogram of Total Steps per Day",
     xlab = "Total Steps",
     ylab = "Frequency",
     col = "skyblue",
     border = "white")

# Add labels and title for mean of total number of steps taken each day
mean_label <- paste("Mean =", round(mean(total_steps_per_day$steps), 2))

# Add labels and title for median of total number of steps taken each day
median_label <- paste("Median =", round(median(total_steps_per_day$steps), 2))

# Add a legend
legend("topright", legend = c(mean_label, median_label), col = c("red", "blue"), bty = "n")


```



# This chunk calculates the Time series plot of the average number of steps taken and The 5-minute interval that, on average, contains the maximum number of steps


```{r average_daily_activity_pattern}
# Group data by 5-minute intervals and calculate average steps
average_steps_by_interval <- aggregate(steps ~ interval, data = activity, mean, na.rm = TRUE)

# Find the interval with the maximum average steps
max_interval <- average_steps_by_interval$interval[which.max(average_steps_by_interval$steps)]
print("The 5-minute interval that, on average, contains the maximum number of steps is:")
print(max_interval)

# Create a time series plot
plot(average_steps_by_interval$interval, average_steps_by_interval$steps, type = "l",
     main = "Average Number of Steps by 5-Minute Interval",
     xlab = "5-Minute Interval",
     ylab = "Average Steps",
     col = "blue")

# Highlight the interval with the maximum average steps
points(max_interval, max(average_steps_by_interval$steps), col = "red", pch = 16, cex = 1.5)

# Add labels and legend
text(max_interval, max(average_steps_by_interval$steps), 
     paste("Max Interval: ", max_interval), pos = 1, col = "red")

legend("topright", legend = c("Average Steps", "Max Interval"), 
       col = c("blue", "red"), lty = 1, pch = c(NA, 16), cex = 0.8)



```


# This chunk executes data imputation. It first replaces the days having all steps as NAs with zeros, and then replaces the days having some NAs with the mean values of that day.


```{r imputation, echo=TRUE}

# Calculate the mean number of steps per day
mean_steps_per_day <- aggregate(steps ~ date, data = activity, mean, na.rm = TRUE)

# Create a bar plot to show the mean number of steps per day
barplot(mean_steps_per_day$steps, 
        names.arg = mean_steps_per_day$date,
        main = "Mean Number of Steps per Day",
        xlab = "Date",
        ylab = "Mean Steps",
        col = "skyblue",
        border = "white",
        las = 2  # Rotates x-axis labels vertically for better readability
)



# Convert the 'date' column to a Date type
activity$date <- as.Date(activity$date)

# Identify days with all steps as NAs for all intervals
days_with_all_na <- activity %>%
  group_by(date) %>%
  filter(all(is.na(steps))) %>%
  pull(date)

# Replace NAs with zeros for days with all steps as NAs
activity <- activity %>%
  mutate(steps = ifelse(date %in% days_with_all_na, 0, steps))

# Calculate mean number of steps per day
mean_steps_per_day <- activity %>%
  group_by(date) %>%
  summarize(mean_steps = mean(steps, na.rm = TRUE))

# create a new dataframe activity1 containing the imputed data
activity1 <- activity %>%
  left_join(mean_steps_per_day, by = "date")

# Replace NAs with mean values for days having only some NAs
activity1 <- activity1 %>%
  mutate(steps = ifelse(is.na(steps) & !date %in% days_with_all_na, mean_steps, steps)) %>%
  select(-mean_steps)

# Display the modified dataset
head(activity1)
```

# This chunk calculates the mean and median of the data after imputation

```{r data_after_imputation, echo=TRUE}


# Calculate the total number of steps taken each day
total_steps_per_day_after_imputation <- aggregate(steps ~ date, data = activity1, sum, na.rm = TRUE)

head(total_steps_per_day_after_imputation)

# Create a histogram
hist(total_steps_per_day_after_imputation$steps, 
     main = "Histogram of Total Steps per Day after imputation",
     xlab = "Total Steps",
     ylab = "Frequency",
     col = "skyblue",
     border = "white")

# Add labels and title for mean of total number of steps taken each day
mean_label <- paste("Mean =", round(mean(total_steps_per_day_after_imputation$steps), 2))

# Add labels and title for median of total number of steps taken each day
median_label <- paste("Median =", round(median(total_steps_per_day_after_imputation$steps), 2))

# Add a legend
legend("topright", legend = c(mean_label, median_label), col = c("red", "blue"), bty = "n")


```


# Finally, this chunk creates a Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r weekdays}
# Create a new column 'day_type' to identify weekdays and weekends
activity$day_type <- ifelse(weekdays(activity$date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

# Calculate the average number of steps per 5-minute interval for weekdays and weekends
average_steps_by_interval <- aggregate(steps ~ interval + day_type, data = activity, mean, na.rm = TRUE)

# Create a panel plot
ggplot(average_steps_by_interval, aes(x = interval, y = steps, group = day_type, color = day_type)) +
  geom_line() +
  facet_wrap(~day_type, scales = "free_y", ncol = 1) +
  labs(title = "Average Number of Steps per 5-Minute Interval",
       x = "5-Minute Interval",
       y = "Average Steps") +
  theme_minimal()


```